Index measures for oak decline severity using phenotypic descriptors

### 2020-08-17 manuscript/manuscript.Rmd
proofing

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].

@@

Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilise a range of oak decline informative phenotypic descriptors observed in oak forests across England.

@@

The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
In each year of sampling surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
The descriptors collected were based on @innes1990 and @icpforests2016.
Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.

@@

The AOD stem symptoms (active bleeds, black staining, calloused wounds) were collected by measuring longest axis of each individual bleed, stain or calloused wound found within the basal 3m of the stem.
The frequency of *A. biguttatus* exit holes were counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.

@@

To account for age, size and site specific tree differences, the tree size descriptors (DBH, lower crown height, timber height, total height, crown radius) were standardised using an adjustment of site means to the overall mean. 
An example of this adjustment is shown for DBH in Supplementary Figure S1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table S2.  
All composite descriptors utilising these adjusted descriptors were recalculated after adjustment.

@@

The remission trees were also found to be significantly different to the AOD (`r glue("t({remission_DAI_group_ttest$parameter[1]}) = {signif(remission_DAI_group_ttest$statistic[1],digits = 3) %>% abs()}, p = < 0.001")`) and COD (`r glue("t({remission_DAI_group_ttest$parameter[2]}) = {signif(remission_DAI_group_ttest$statistic[2],digits = 3) %>% abs()}, p = {round(remission_DAI_group_ttest$p.value[2],3)}")`) groups on the DAI scale with group means of `r signif(remission_DAI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_DAI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_DAI_group_ttest$estimate1[2],digits = 3)` for the AOD, remission and COD groups respectively. 
As a group, the remission trees were found to be neither AOD or COD.

@@

Plots of the individual descriptors against the decline indexes are available in Supplementary Figures S5-S8. 
A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_PDI$ and $M_DAI$ models.
The following sections (3.4, 3.5) identify how the descriptors contribute and interact within the decline index models in specific decline cases.

@@

PDI_lime_analysis_table %>%
  kable(format = 'latex',
        booktabs = TRUE,
        escape = FALSE,
        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$. Top four important descriptors show. Descriptors ordered by descending absolute weight within each case.') %>%
  column_spec(1,width = '2cm') %>%
  column_spec(3,width = '2cm') %>%
  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
```

@@

Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
This is to be expected and are indicators of model suitability using the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure \@ref(descriptor-contributions)a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure \@ref(fig:PDI-response-surfaces)a).

>>>>

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of callused tissue on the exterior [@denman2014].
Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].

@@

Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilise a range of oak decline informative phenotypic descriptors, observed in oak forests across England.

@@

The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
In each year of sampling, surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
The descriptors collected were based on @innes1990 and @icpforests2016.
Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.

@@

The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
The frequency of *A. biguttatus* exit holes were counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.

@@

To account for age, size and site specific tree differences, the tree size descriptors (DBH, lower crown height, timber height, total height, crown radius) were standardised using an adjustment of site means to the overall mean. 
An example of this adjustment is shown for DBH in Supplementary Figure S1 and the adjustment factors for each of the descriptors at each site are given in Supplementary Table S2.  
All composite descriptors utilising these adjusted descriptors were recalculated after adjustment.

@@

The remission trees were also found to be significantly different to the AOD (`r glue("t({remission_DAI_group_ttest$parameter[1]}) = {signif(remission_DAI_group_ttest$statistic[1],digits = 3) %>% abs()}, p = < 0.001")`) and COD (`r glue("t({remission_DAI_group_ttest$parameter[2]}) = {signif(remission_DAI_group_ttest$statistic[2],digits = 3) %>% abs()}, p = {round(remission_DAI_group_ttest$p.value[2],3)}")`) groups on the DAI scale with group means of `r signif(remission_DAI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_DAI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_DAI_group_ttest$estimate1[2],digits = 3)` for the AOD, remission and COD groups respectively. 
As a group, the remission trees were found to be neither distinctively AOD or COD.

@@

Plots of the individual descriptors against the decline indexes are available in Supplementary Figures S5-S8. 
A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_{PDI}$ and $M_{DAI}$ models.
The following sections (3.4, 3.5) identify how the descriptors contribute and interact within the decline index models in specific decline cases.

@@

PDI_lime_analysis_table %>%
  kable(format = 'latex',
        booktabs = TRUE,
        escape = FALSE,
        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$. Top four important descriptors shown. Descriptors ordered by descending absolute weight within each case.') %>%
  column_spec(1,width = '2cm') %>%
  column_spec(3,width = '2cm') %>%
  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
```

@@

Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
This is to be expected and are indicators of model suitability using the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007]. 
However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure \@ref(fig:descriptor-contributions)a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure \@ref(fig:PDI-response-surfaces)a).
